---
title: "Examples and recipes"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
library(zumsgpack)
```

## Nanosecond timestamps

A `POSIXct` is a double, which near 2026 resolves about 2^-22 seconds, so
a timestamp's nanoseconds survive only approximately. When they matter,
a handler for type `"-1"` keeps the fields exactly. The payload is one of
three layouts:

```{r}
timestamp_fields <- function(data) {
  be <- function(b) sum(as.numeric(b) * 256^(rev(seq_along(b)) - 1))
  switch(as.character(length(data)),
    "4"  = c(seconds = be(data), nanoseconds = 0),
    "8"  = {
      hi <- be(data[1:4])
      c(seconds = (hi %% 4) * 2^32 + be(data[5:8]), nanoseconds = hi %/% 4)
    },
    "12" = c(seconds = msgpack_decode(c(as.raw(0xd3), data[5:12])),
             nanoseconds = be(data[1:4])))
}

b <- as.raw(c(0xd7, 0xff, 0xa1, 0xdc, 0xd7, 0xc8, 0x5a, 0x4a, 0xf6, 0xa5))
msgpack_decode(b)                                        # to double precision
msgpack_decode(b, ext_handlers = list("-1" = timestamp_fields))
```

## A type of your own

An extension type carries bytes under a number from 0 to 127 that both
sides agree on. `as_msgpack()` writes it; `ext_handlers` reads it back.
Here, UUIDs as type 37:

```{r}
uuid <- function(hex) structure(as.raw(strtoi(substring(gsub("-", "", hex),
                                seq(1, 31, 2), seq(2, 32, 2)), 16L)), class = "uuid")
format.uuid <- function(x, ...) {
  h <- paste(sprintf("%02x", as.integer(unclass(x))), collapse = "")
  paste(substring(h, c(1, 9, 13, 17, 21), c(8, 12, 16, 20, 32)), collapse = "-")
}
print.uuid <- function(x, ...) cat("<uuid>", format(x), "\n")
as_msgpack.uuid <- function(x, ...) msgpack_ext(37, unclass(x))
registerS3method("as_msgpack", "uuid", as_msgpack.uuid)

id <- uuid("6ba7b810-9dad-11d1-80b4-00c04fd430c8")
b <- msgpack_encode(list(id = id, n = 1L))
msgpack_annotate(b)
msgpack_decode(b, ext_handlers = list("37" = function(data) structure(data, class = "uuid")))
```

## A Fluentd forward stream

Fluentd's forward protocol sends `[tag, [[time, record], ...]]` messages
back to back, with the time as an `EventTime` extension (type 0: 32-bit
seconds, then 32-bit nanoseconds). `msgpack_read_seq(each =)` reads one
message at a time:

```{r}
event_time <- function(s, ns = 0) {
  be32 <- function(v) as.raw(c(v %/% 2^24, (v %/% 2^16) %% 256, (v %/% 256) %% 256, v %% 256))
  msgpack_ext(0, c(be32(s), be32(ns)))
}
wire <- msgpack_encode_seq(list(
  list("app.access", list(list(event_time(1.7e9, 5e8), list(path = "/", status = 200L)),
                          list(event_time(1.7e9 + 1), list(path = "/x", status = 404L)))),
  list("app.error", list(list(event_time(1.7e9 + 2), list(message = "oops"))))
))

as_time <- function(d) {
  be <- function(b) sum(as.numeric(b) * 256^(3:0))
  as.POSIXct(be(d[1:4]) + be(d[5:8]) / 1e9, origin = "1970-01-01", tz = "UTC")
}
con <- rawConnection(wire)
msgpack_read_seq(con, ext_handlers = list("0" = as_time), each = function(m) {
  for (entry in m[[2]]) {
    what <- if (is.null(entry[[2]]$path)) entry[[2]]$message else entry[[2]]$path
    cat(m[[1]], format(entry[[1]], "%H:%M:%OS1"), what, "\n")
  }
})
close(con)
```

## Tables

An array of records -- one map per row, the way most services send a
table -- becomes a data frame with `data_frame = TRUE`. A key a row lacks
is `NA`, and a column of mixed kinds is a list column:

```{r}
rows <- msgpack_encode(list(
  list(id = 1L, name = "a", score = 1.5),
  list(id = 2L, name = "b"),
  list(id = 3L, score = NULL, tags = c("x", "y"))
))
msgpack_decode(rows, data_frame = TRUE)
```

## MessagePack or CBOR

zumsgpack follows zucbor's design: the same check-then-build decoder, the
same limits, the same mapping to R and the same deterministic encoder. A
protocol that can speak either changes a prefix:

```{r, eval = requireNamespace("zucbor", quietly = TRUE)}
x <- list(alg = -7L, kid = "11", when = as.POSIXct(0, tz = "UTC"))
str(zucbor::cbor_decode(zucbor::cbor_encode(x)))
str(msgpack_decode(msgpack_encode(x)))
```
