---
title: "Usage examples"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This article walks through zumsgpack's functions with small, runnable
examples: encoding and decoding, how R values map to MessagePack and back,
timestamps and extension types, files and streams, tables, and what
happens when the input is bad.

```{r}
library(zumsgpack)
```

## Encode and decode

`msgpack_encode()` turns an R value into a raw vector of MessagePack bytes,
and `msgpack_decode()` turns bytes back into an R value.

```{r}
b <- msgpack_encode(list(id = 7L, name = "sensor", values = c(1.5, 2.25)))
b
length(b)
str(msgpack_decode(b))
```

Decoding needs exactly one object; `msgpack_decode_seq()` reads zero or
more objects written back to back, and `msgpack_encode_seq()` writes them.

```{r}
seq_bytes <- msgpack_encode_seq(list(1L, "two", TRUE))
seq_bytes
msgpack_decode_seq(seq_bytes)
```

## Vectors, lists and missing values

An atomic vector is an array; a length-one vector is a single value unless
it is wrapped in `I()` or `auto_unbox = FALSE`.

```{r}
msgpack_encode(1:3)
msgpack_encode(5L)
msgpack_encode(I(5L))
msgpack_encode(5L, auto_unbox = FALSE)
```

On the way back, an array becomes an atomic vector when its elements agree
and a list otherwise. A one-element array comes back marked with `I()`, so
that encoding it again gives the same bytes.

```{r}
msgpack_decode(msgpack_encode(list(1L, 2.5)))     # numbers combine
msgpack_decode(msgpack_encode(list(1L, "a")))     # mixed kinds: a list
str(msgpack_decode(msgpack_encode(I(5L))))     # an array of one, marked I()
```

`NULL` and `NA` are both `nil`. Inside a vector, `nil` comes back as `NA`;
on its own, as `NULL`.

```{r}
msgpack_encode(c(1L, NA, 3L))
msgpack_decode(msgpack_encode(c(1L, NA, 3L)))
msgpack_decode(msgpack_encode(NA))
```

Use `simplify = "none"` to get every array as a list.

```{r}
msgpack_decode(msgpack_encode(1:3), simplify = "none")
```

## Maps

A fully named list or vector is a map with string keys, and a map whose
keys are all non-empty, unique strings comes back as a named list.

```{r}
msgpack_decode(msgpack_encode(c(a = 1L, b = 2L)))
```

Map entries are written sorted by their encoded keys (shorter keys first),
so the same values always give the same bytes, whatever order the list
was built in:

```{r}
identical(msgpack_encode(list(b = 1, a = 2)), msgpack_encode(list(a = 2, b = 1)))
```

Maps with other keys -- integers, binary, nested values -- are
`msgpack_map` objects, which keep the keys as they are. Build one with
`msgpack_map()`:

```{r}
m <- msgpack_map(list(1L, 2L), list("one", "two"))
b <- msgpack_encode(m)
msgpack_decode(b)
```

`map_keys` changes how maps are decoded: `"map"` always gives a
`msgpack_map`, and `"string"` always gives a named list, naming non-string
keys by a text rendering of them.

```{r}
msgpack_decode(b, map_keys = "string")
msgpack_decode(msgpack_encode(list(a = 1L)), map_keys = "map")
```

## Numbers

Integers come back as `integer` when R's integer type holds them, as
`double` up to 2^53, and beyond that as `msgpack_bigint`, which keeps the
exact value as decimal text. `big_integers` chooses otherwise.

```{r}
big <- msgpack_encode(msgpack_bigint("18446744073709551615"))   # 2^64 - 1
msgpack_decode(big)
msgpack_decode(big, big_integers = "double")
```

A whole double is written as an integer, since R has no integer literal:
`1` and `1L` encode the same. Other doubles are `float 64`; with
`floats = "shortest"`, a `float 32` when it holds the value exactly.

```{r}
msgpack_encode(1)
msgpack_encode(0.5)
msgpack_encode(0.5, floats = "shortest")
```

## Strings and bytes

Character vectors are UTF-8 `str`; raw vectors are `bin`.

```{r}
msgpack_encode("héllo")
msgpack_decode(msgpack_encode(as.raw(c(0xde, 0xad, 0xbe, 0xef))))
```

## Timestamps and dates

`POSIXct` is written as MessagePack's timestamp extension, in the smallest
of its three layouts, and comes back as a `POSIXct` in UTC. A `Date` is a
timestamp at midnight UTC.

```{r}
when <- as.POSIXct("2026-10-09 12:30:00.25", tz = "UTC")
b <- msgpack_encode(when)
b
format(msgpack_decode(b), "%Y-%m-%d %H:%M:%OS3 %Z")
msgpack_decode(msgpack_encode(as.Date("2026-10-09")))
```

`ext = "keep"` leaves timestamps as raw extension objects.

```{r}
msgpack_decode(b, ext = "keep")
```

## Extension types

An extension object is a type number and some bytes. Types without a
conversion come back as `msgpack_ext`, and `msgpack_ext()` builds one.

```{r}
e <- msgpack_ext(5, charToRaw("hi"))
msgpack_decode(msgpack_encode(e))
```

`ext_handlers` gives a type meaning on decode: each handler gets the
payload as a raw vector and returns any R value.

```{r}
msgpack_decode(msgpack_encode(list(e, e)), ext_handlers = list("5" = rawToChar))
```

The encoding half is the `as_msgpack()` generic: a method for your class
returns something zumsgpack can encode, often a `msgpack_ext()`.

```{r}
point <- function(x, y) structure(list(x = x, y = y), class = "point")
as_msgpack.point <- function(x, ...) {
  msgpack_ext(7, writeBin(c(x$x, x$y), raw(), size = 4, endian = "big"))
}
registerS3method("as_msgpack", "point", as_msgpack.point)

# "tip" is decoded before "origin": map keys are written sorted, shorter first.
b <- msgpack_encode(list(origin = point(0, 0), tip = point(3, 4)))
str(msgpack_decode(b, ext_handlers = list("7" = function(data) {
  v <- readBin(data, "double", n = 2, size = 4, endian = "big")
  point(v[1], v[2])
})))
```

## Files and connections

`msgpack_read()` reads one object from a path, URL or connection, and
reads at most `max_size + 1` bytes, so an oversized file fails rather than
filling memory.

```{r}
path <- tempfile(fileext = ".msgpack")
writeBin(msgpack_encode(list(a = 1L, b = "x")), path)
msgpack_read(path)
```

`msgpack_read_seq()` reads objects written back to back. Without `each`
it returns a list; with `each`, every object is checked whole, decoded and
passed to your function as soon as it arrives, so memory is bounded by the
largest object rather than the whole stream.

```{r}
writeBin(msgpack_encode_seq(lapply(1:5, function(i) list(seq = i, ok = i %% 2 == 1))), path)
msgpack_read_seq(path)[[2]]

ok <- 0
n <- msgpack_read_seq(path, each = function(m) if (m$ok) ok <<- ok + 1)
c(read = n, ok = ok)
unlink(path)
```

## The object at the start of a buffer

When MessagePack sits inside other framing, `msgpack_decode_prefix()`
decodes the first object and reports how many bytes it used, without
reading what follows.

```{r}
x <- c(msgpack_encode(1:2), as.raw(c(0xde, 0xad, 0xbe, 0xef)))
r <- msgpack_decode_prefix(x)
r$value
x[-seq_len(r$consumed)]
```

## Tables

`data_frame = TRUE` turns an array of records -- one map per row -- into a
data frame. A key a row lacks is `NA`, and a column of mixed kinds is a
list column. A data frame encodes as such an array.

```{r}
df <- data.frame(id = 1:3, name = c("a", "b", NA), score = c(1.5, NA, 3))
b <- msgpack_encode(df)
msgpack_decode(b, data_frame = TRUE)
```

Columns come back in sorted key order, since map keys are sorted on
encode.

## Bad input

Input is checked whole before anything is built. `msgpack_validate()` runs
just that check and says whether it passed.

```{r}
msgpack_validate(msgpack_encode(1:3))
msgpack_validate(as.raw(c(0x93, 0x01, 0x02)))   # promises 3 elements, has 2
```

Every error is a classed condition inheriting `zumsgpack_error`, with the
byte offset of the fault, so it can be caught by kind.

```{r}
tryCatch(
  msgpack_decode(as.raw(c(0x82, 0x01, 0xc0, 0x01, 0xc0))),   # {1: nil, 1: nil}
  zumsgpack_duplicate_key = function(e) conditionMessage(e)
)
```

Limits bound what an input can cost: `max_size` (bytes), `max_depth`
(nesting), `max_items` (objects) and, for data frames, `max_cells`.

```{r}
deep <- c(rep(as.raw(0x91), 300), as.raw(0xc0))   # 300 nested arrays
tryCatch(msgpack_decode(deep), zumsgpack_depth_limit = function(e) e$limit_value)
msgpack_decode(deep, max_depth = 300L) |> class()
```

See `vignette("untrusted")` for more on decoding input you do not control.

## Looking at the bytes

`msgpack_annotate()` prints an annotated hex dump: each head's offset, its
bytes, and what it is, indented by depth.

```{r}
msgpack_annotate(msgpack_encode(list(id = 7L, tags = c("a", "b"), when = when)))
```

## Build information

```{r}
zumsgpack_info()
```
