---
title: "Getting started with zumsgpack"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
library(zumsgpack)
```

## Encoding and decoding

`msgpack_encode()` turns an R value into MessagePack bytes and
`msgpack_decode()` turns them back:

```{r}
x <- list(name = "sensor-1", values = c(20.5, 21.25), ok = TRUE)
b <- msgpack_encode(x)
b
str(msgpack_decode(b))
```

Arrays whose elements agree become atomic vectors; anything else is a
list. Maps with string keys become named lists; maps with other keys
become a `msgpack_map`, which keeps the keys as they were:

```{r}
msgpack_decode(msgpack_encode(msgpack_map(list(1L, 2L), list("one", "two"))))
```

## The same bytes, every time

Encoding is deterministic: the smallest integer forms, sorted map keys,
one form for every value. Identical R values give identical bytes on every
platform, which a hash or a signature over the bytes needs:

```{r}
identical(msgpack_encode(list(b = 1, a = 2)), msgpack_encode(list(a = 2, b = 1)))
```

## Timestamps and extension types

The timestamp extension becomes a `POSIXct` in UTC, and back:

```{r}
when <- as.POSIXct("2026-10-09 12:00:00.25", tz = "UTC")
msgpack_decode(msgpack_encode(when))
```

Other extension types come back as `msgpack_ext`, unless you give them a
handler; see `vignette("untrusted")` for decoding input you do not
control, and the examples article on the package site for recipes.

## Files, streams and tables

```{r}
path <- tempfile(fileext = ".msgpack")
writeBin(msgpack_encode(list(list(id = 1L, v = "a"), list(id = 2L, v = "b"))), path)
msgpack_read(path, data_frame = TRUE)
unlink(path)
```
